### 2010-03-29 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: Tidied up the code in the paper, fixed a few transcription bugs and added colour (well, purple)

While inductive types are the alpha and omega of simply-typed
languages, evolving in a dependent setting fosters new
opportunities. The typical example is bounded lists, also known as
vectors. A vector is a list decorated by its length. Having this
information prevents hazardous operations, such as taking the head of
an empty vector: the head function only takes vectors of length
$\NatSuc{n}$, as enforced by its type. This is made possible by the
very specificity of dependently-typed systems: terms -- the length --
are allowed to influence types -- the vector type. \note{Reference
  paper to cite about vectors in dependent types?}

However, this new class of objects cannot be defined by mere
induction. In the case of vectors, for instance, we have to define the
whole \emph{family} of vector in one go: vectors of all size needs to
be defined at the same time. Where simply-typed languages have
inductive types, in dependently typed programming we the basic grammar of
data-types should be that of inductive families of types. To this end, we
rely on \emph{indexing}. We are going to transform the $\Desc$ universe into
$\IDesc$, the universe of indexed descriptions. We express inductive families
by the $I \To \IDesc{I}$ type.  \note{pwm: Vec might not be the best
motivating example here, since it doesn't appear in the paper (more on that
later). Come to think of it, this is entirely too vague without spelling out
a concrete example.}

@@

$$\Bhab{\Sigma}{\PI{\V{S}}{\Set} (S \To \Set) \To \Set}$$

In the alternative presentation, we go further and present all our
codes as quotations of standard type-formers. This presentation is
shown in Figure~\ref{fig:type-former-desc}. The reader will notice
that we replace $\DUnit$ by a more general $\DConst$ code. Whereas
$\DUnit$ was interpreted as the unit set, $\DConst{X}$ is interpreted
as $X$, for any $\Bhab{X}{\Set}$. Extensionally, $\DConst{X}$ and
$\DSigma{X}{\DUnit}$ are equivalent. However, $\DConst$ is more
succinct. More importantly, $\DConst$ is \emph{first-order}, unlike
its equivalent encoding. From a definitional perspective, we are
giving more opportunities to the type-system, hence reducing the
burden on the programmer. For the same reason, we introduce
$\DProd{}{}$ that overlaps with $\DSigma{}{}$.

@@

\[\begin{array}{l@{\:\mapsto\:\:}l}
\DIndx{D}         & \DProd{\DId}{D}                      \\
\DHindx{H}{D}     & \DProd{(\DPi{H}{(\LAM{\_} \DId)})}{D}
\end{array}
\]

@@

\[\stk{
\begin{array}{ll}
\stk{
\data \Desc : \Set \where                                      \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \DId            & \Desc                                    \\
    \DConst{}       & \Set \To \Desc                           \\
    \DProd{}{}      & \PI{\V{D}, \V{D'}}{\Desc} \Desc          \\
    \DSigma         & \PI{\V{S}}{\Set} \PIS{S \To \Desc} \Desc \\
    \DPi            & \PI{\V{S}}{\Set} \PIS{S \To \Desc} \Desc 
\end{array}
}
\vspace{0.2in}
\\
\stk{
\descop{\_\:}{} : \Desc \To \Set \To \Set \\
\begin{array}{@{}l@{\:=\:\:}ll}
\descop{\DId}{X}          &  X                                           \\
\descop{\DConst{Z}}{X}    &  Z                                           \\
\descop{\DProd{D}{D'}}{X} &  \TIMES{\descop{D}{X}}{\descop{D'}{X}}       \\
\descop{\DSigma{S}{D}}{X} &  \SIGMA{\V{s}}{S} \descop{D\: s}{X}          \\
\descop{\DPi{S}{D}}{X}    &  \PI{\V{s}}{S} \descop{D\: s}{X}            
\end{array}
}
\end{array}
}\]

@@

In the previous section, we have presented the $\Desc$ universe as the
grammar of functors in the category $\Set$. We have seen how to code
inductive types in this setting. To describe an inductive family
indexed by $\Bhab{I}{\Set}$, we use endofunctors on the category
$\Set^I$. We call these \emph{indexed functors}. \note{Predicate
  transformer anyone?} In this setting, $I \To \IDesc{I}$ is our
grammar for describing these functors. Hence, $\IDesc{}$ and its
interpretation have the following types:

\[\stk{
\data \IDesc{} (\Bhab{\V{I}}{\Set}) : \Set \where \\
\;\;\ldots \\
\\
\idescop{\_\:}{}{} : \PI{\V{I}}{\Set} \IDesc{I} \To (\V{I} \To \Set) \To \Set    \\
\ldots
}\]

Note that $I \To \IDesc{I}$ is isomorphic, up to interpretation, to an
endofunctor on $\Set^I$. Indeed, $I \To \IDesc{I}$ is interpreted as
$I \To (I \To \Set) \To \Set$, which is isomorphic to $(I \To \Set)
\To (I \To \Set)$. Fix-point of inductive families is defined over
these indexed functors, hence computing a fix-point of the entire
\emph{family} of functors:

\[\stkl{
\Rule{\stk{ 
          \Gamma \vdash \Bhab{I}{\Set} \\
          \Gamma \vdash \Bhab{R}{I \To \IDesc{I}} \\
          \Gamma \vdash \Bhab{i}{I}}}
     {\Gamma \vdash \Bhab{\IMu{I}{R}{i}}{\Set}} \qquad
\Rule{\stk{
          \Gamma \vdash \Bhab{I}{\Set} \\
          \Gamma \vdash \Bhab{R}{I \To \IDesc{I}} \\
          \Gamma \vdash \Bhab{i}{I} \\
          \Gamma \vdash \Bhab{x}{\idescop{R\: i}{I}{(\IMu{I}{R}{})}}}}
     {\Gamma \vdash \Bhab{\Con{x}}{\IMu{I}{R}{i}}}
}\]

@@

\[\stk{
\begin{array}{ll}
\stk{
\data \IDesc{} (\Bhab{\V{I}}{\Set}) : \Set \where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \DVar{}         & I \To \IDesc{I}                                   \\
    \DConst{}       & \Set \To \IDesc{I}                                \\
    \DProd{}{}      & \PI{\V{D}, \V{D'}}{\IDesc{I}} \IDesc{I}           \\
    \DSigma         & \PI{\V{S}}{\Set} \PIS{S \To \IDesc{I}} \IDesc{I}  \\
    \DPi            & \PI{\V{S}}{\Set} \PIS{S \To \IDesc{I}} \IDesc{I} 
\end{array}
}
\vspace{0.2in}
\\
\stk{
\idescop{\_\:}{}{} :_{\PI{\V{I}}{\Set}} \IDesc{I} \To (\V{I} \To \Set) \To \Set                  \\
\begin{array}{@{}l@{\:=\:\:}ll}
\idescop{\DVar{i}}{I}{P}      &  P~i                                                 \\
\idescop{\DConst{X}}{I}{P}    &  X                                                   \\
\idescop{\DProd{D}{D'}}{I}{P} &  \TIMES{\idescop{D}{I}{P}}{\idescop{D'}{I}{P}}       \\
\idescop{\DSigma{S}{D}}{I}{P} &  \SIGMA{\V{s}}{S} \idescop{D\: s}{I}{P}                    \\
\idescop{\DPi{S}{D}}{I}{P}    &  \PI{\V{s}}{S} \idescop{D\: s}{I}{P}            
\end{array}
}
\end{array}
}\]

@@

\[\stk{
 \TagIDesc{I}  \mapsto \TIMES{\ATagIDesc{I}}{\ITagIDesc{I}} \\
 \ATagIDesc{I} \mapsto \SIGMA{\V{E}}{\EnumU} (\PI{\V{i}}{I} \spi{\V{E}}{(\LAM{\_} \IDesc{I})}) \\
 \ITagIDesc{I} \mapsto \SIGMA{\V{F}}{I \To \EnumU} (\PI{\V{i}}{I} \spi{(F\: i)}{(\LAM{\_} \IDesc{I})}) 
}\]

@@

\[\stk{
\VecD : \Set \To \TagIDesc{\Nat} \\
\begin{array}{@{}ll}
\VecD\: X = \bigRedBracket{ 
            \begin{array}{l}
                \pair{[]}{\_}{} \\
                \bigRedBracket{
                \begin{array}{l} 
                  \LAM{n} \caseB{n}{\NatZero \To [\VecNil] \\
                                   (\NatSuc n) \To [\VecCons{}{}]} \\
                  \LAM{n} \caseB{n}{\NatZero \To [\DConst{Unit}] \\
                                   (\NatSuc n) \To [\DSigma{X}{\DVar{n}}]}
                \end{array}}
            \end{array}}
\end{array}
}\]

@@

\[\stk{
\IDescD : \PI{\V{I}}{\Set} \IDesc{\Unit} \\
\begin{array}{@{}ll}
\IDescD~I \mapsto \DSigma{}{} & (\EnumT \red{[} \DVar{},
                                          \DConst{},
                                          \DProd{}{},
                                          \DSigma{}{}, 
                                          \DPi{}{} \red{]}) \\
                              & \bigRedBracket{\begin{array}{l}
                                      \DConst{I}                  \\
                                      \DConst{\Set}               \\
                                      \DProd{\DVar{\Void}}{\DVar{\Void}}  \\
                                      \DSigma{\Set}{(\LAM{S} \DPi{S}{(\LAM{\_} \DVar{\Void})})} \\
                                      \DSigma{\Set}{(\LAM{S} \DPi{S}{(\LAM{\_} \DVar{\Void})})}
                                   \end{array}
                             }
\end{array}
}\]

@@

\[
\begin{array}{l@{\::\:\:}l}
\Econd{}{}{}     & \forall \Bhab{ty}{\Ty} . \Ebool \To ty \To ty \To ty  \\ 
\Eplus{}{}       & \Enat \To \Enat \To \Enat                           \\
\Ele{}{}         & \Enat \To \Enat \To \Ebool                          \\
\Eval{}          & \forall \Bhab{ty}{\Ty} . \Val{ty} \To ty
\end{array}
\]

@@

\[\stkc{
\stk{
\HExprD : \TagIDesc{\Ty} \\
\HExprD \mapsto ( \HExprAD , \HExprID ) \\
} \\
\\
\begin{array}{ll}
\stk{
\HExprAD : \ATagIDesc{\Ty} \\
\HExprAD \mapsto \bigRedBracket{
                 \begin{array}{l}
                   \EnumT{[\Eval{}, \Econd{}{}{}]} \red{,} \\
                   \LAM{ty}
                   \bigRedBracket{
                   \begin{array}{l}
                   \DConst{(\Val{ty})} \\
                   \DProd{\DVar{\Ebool}}{\DProd{\DVar{ty}}{\DVar{ty}}}
                   \end{array}
                   }
                 \end{array}
                 }
\\
\\
} &
\stk{
\HExprID : \ITagIDesc{\Ty} \\
\HExprID \mapsto \bigRedBracket{
                 \begin{array}{l}
                   \LAM{ty} \case{ty}{\Enat \To \EnumT{[\Eplus{}{}]} \\ \Ebool \To \EnumT{[\Ele{}{}]}} \red{,} \\
                   \LAM{ty}
                   \case{ty}{
                     \Enat \To \DProd{\DVar{nat}}{\DVar{nat}} \\
                     \Ebool \To \DProd{\DVar{nat}}{\DVar{nat}}
                   }
                   \end{array}
                   }
}
\end{array}
}\]

@@

\[
\evalH : \PI{\V{ty}}{\Ty} 
         \IMu{\Ty}{\toIDesc{\HExprD}}{ty} \To
         \Val{ty}
\]

@@

\[\stk{
\evalOne : \PI{\V{ty}}{\Ty} \idescop{\toIDesc{\HExprD}\: ty}{\Ty}{ty} \To {\Val{ty}} \\
\begin{array}{@{}l@{\:=\:\:}l}
\evalOne\: \_\: \pair{\Eval{}}{x}{}                                         & x \\
\evalOne\: \_\: \pair{\Econd{}{}{}}{\pair{\BoolTrue}{\pair{x}{\_}{}}{}}{}   & x \\
\evalOne\: \_\: \pair{\Econd{}{}{}}{\pair{\BoolFalse}{\pair{\_}{y}{}}{}}{}  & x \\
\evalOne\: \Enat\: \pair{\Eplus{}{}}{\pair{x}{y}{}}{}                       & \F{plusHost}\: x\: y \\
\evalOne\: \Ebool\: \pair{\Ele{}{}}{\pair{x}{y}{}}{}                        & \F{leHost}\: x\: y 
\end{array} \\
\\
\evalH : \PI{\V{ty}}{\Ty} 
           \IMu{\Ty}{\toIDesc{\HExprD}}{ty} \To
           \Val{ty} \\
\evalH\: ty\: term = \F{cataI}\: \Ty\: 
                                 \toIDesc{\HExprD}\: 
                                 \Val{}\: 
                                 \evalOne\: 
                                 ty\: 
                                 term
}\]
\note{pwm: $\evalOne$ rather illustrates my point about the 2 types of
  IFunc floating around, $\idescop{\_}{}{}$ is defined using the 1st
  notion, and used it here as if it were the 2nd. pierre: in this
  case, it was a typo. But you're right, there is two notions and we
  probably want to clarify where these things live.}

@@

\[\stk{
\Var{}{} : \EnumU \To \Ty \To \Set \\
\Var{dom}{\_} = \EnumT{dom}
}\]

Whereas $\Val{}$ was mapping the type to the corresponding host type,
$\Var{}{}$ maps types to a finite set. The finite set -- the context
-- contains closed terms. A variable is therefore a $\Eval{}$ that
contains a pointer to a particular element of the finite set -- an
element of $\EnumT{dom}$. The extra argument to $\Var{}{}$ is the
domain of this context.

Consequently, replacing $\Val{ty}$ by
$(\SUM{\Val{ty}}{\Var{dom}{ty}})$ in Figure~\ref{fig:hexpr-full} turns
the language of closed term into a language of opened terms with
constants. For readability, we will abbreviate $\LAM{ty}
\SUM{\Val{ty}}{\Var{dom}{ty}}$ into $\SUM{\Val{}}{\Var{}{}}$ \note{Is
  it polite and comprehensible to ask that?} This defines a new
indexed description, called $\HExprVarD{dom}$.

@@

\[\stk{
\begin{array}{@{}ll}
\discharge : & \PITEL{\V{ty}}{\Ty}
               \PITEL{\V{dom}}{\EnumU} \\
             & \PI{\V{\gamma}}{\spi{dom}{(\LAM{\_} \IMu{\Ty}{\toIDesc{\HExprVarD{dom}}}{ty})}} \\
             & (\SUM{\Val{ty}}{\Var{dom}{ty}}) \To
               \IMu{\Ty}{\toIDesc{\HExprD}}{ty} 
\end{array} \\
\begin{array}{@{}l@{\:\mapsto\:\:}l}
\discharge\: ty\: dom\: \gamma\: (\SumLeft x)  & \Con{\pair{\Eval{}}{x}{}} \\
\discharge\: ty\: dom\: \gamma\: (\SumRight v) & \switch{dom}{(\LAM{\_} \IMu{\Ty}{\toIDesc{\HExprVarD{dom}}}{ty})}{\gamma}{v}
\end{array}
}\]

@@

\newcommand{\substH}{\F{substExpr}}
\newcommand{\domNat}{dom_{\CN{nat}}}
\newcommand{\domBool}{dom_{\CN{bool}}}
\newcommand{\gammaNat}{\gamma_{\CN{nat}}}
\newcommand{\gammaBool}{\gamma_{\CN{bool}}}
\newcommand{\GammaSpi}[2]{\Gamma_{\CN{ty}}~#1~#2}

\[
\begin{array}{@{}ll}
\substH  : & \PITEL{\V{\dom}}{\EnumU} \\
           & \PITEL{\V{\gammaNat}}{\GammaSpi{dom}{\Enat}} 
             \PITEL{\V{\gammaBool}}{\GammaSpi{dom}{\Ebool}} \\
           & \begin{array}{@{}ll}
             \PI{\V{\sigma}}{& \PITEL{\V{dom}}{\EnumU} \\
                             & \PITEL{\V{\gammaNat}}
                                     {\GammaSpi{dom}{\Enat}} \\
                             & \PITEL{\V{\gammaBool}}
                                     {\GammaSpi{dom}{\Ebool}} \\
                             & \PI{\V{ty}}{\Ty} (\SUM{\Val{ty}}{\Var{dom}{ty}}) \To \\
                             & \IMu{\Ty}{(\toIDesc{\HExprD})}{ty}}
             \end{array}\\
          & \PI{\V{ty}}{\Ty}
            \IMu{\Ty}{\toIDesc{\HExprVarD{dom}}}{\V{ty}} \To \\
          & \IMu{\Ty}{\toIDesc{\HExprD}}{\V{ty}}
\end{array}
\]

@@

\[
\sigma\: dom\: \gammaNat\: \gammaBool\: ty\: var \mapsto 
    \discharge\: ty\: dom\: \gamma_{ty}\: var 
\]

>>>>

While inductive types are the alpha and omega of simply-typed
languages, evolving in a dependent setting fosters new
opportunities. The typical example is bounded lists, also known as
vectors. A vector is a list decorated by its length. Having this
information prevents hazardous operations, such as taking the head of
an empty vector: the head function only takes vectors of length
$\NatSuc{\V{n}}$, as enforced by its type. This is made possible by the
very specificity of dependently-typed systems: terms -- the length --
are allowed to influence types -- the vector type. \note{Reference
  paper to cite about vectors in dependent types?}

However, this new class of objects cannot be defined by mere
induction. In the case of vectors, for instance, we have to define the
whole \emph{family} of vector in one go: vectors of all size needs to
be defined at the same time. Where simply-typed languages have
inductive types, in dependently typed programming we the basic grammar of
data-types should be that of inductive families of types. To this end, we
rely on \emph{indexing}. We are going to transform the $\Desc$ universe into
$\IDesc$, the universe of indexed descriptions. We express inductive families
by the $\V{I} \To \IDesc{\V{I}}$ type.  \note{pwm: Vec might not be the best
motivating example here, since it doesn't appear in the paper (more on that
later). Come to think of it, this is entirely too vague without spelling out
a concrete example.}

@@

$$\Bhab{\Sigma}{\PI{\V{S}}{\Set} \PI{\V{S}}{\Set} \Set}$$

In the alternative presentation, we go further and present all our
codes as quotations of standard type-formers. This presentation is
shown in Figure~\ref{fig:type-former-desc}. The reader will notice
that we replace $\DUnit$ by a more general $\DConst$ code. Whereas
$\DUnit$ was interpreted as the unit set, $\DConst{\V{X}}$ is interpreted
as $\V{X}$, for any $\Bhab{\V{X}}{\Set}$. Extensionally, $\DConst{\V{X}}$ and
$\DSigma{\V{X}}{\DUnit}$ are equivalent. However, $\DConst$ is more
succinct. More importantly, $\DConst$ is \emph{first-order}, unlike
its equivalent encoding. From a definitional perspective, we are
giving more opportunities to the type-system, hence reducing the
burden on the programmer. For the same reason, we introduce
$\DProd{}{}$ that overlaps with $\DSigma{}{}$.

@@

\[\begin{array}{l@{\:\mapsto\:\:}l}
\DIndx{\V{D}}         & \DProd{\DId}{\V{D}}                      \\
\DHindx{\V{H}}{\V{D}}     & \DProd{(\DPi{\V{H}}{(\LAM{\_} \DId)})}{\V{D}}
\end{array}
\]

@@

\[\stk{
\begin{array}{ll}
\stk{
\data \Desc : \Set \where                                      \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \DId            & \Desc                                    \\
    \DConst{}       & \Set \To \Desc                           \\
    \DProd{}{}      & \PI{\V{D}, \V{D'}}{\Desc} \Desc          \\
    \DSigma         & \PI{\V{S}}{\Set} \PIS{\V{S} \To \Desc} \Desc \\
    \DPi            & \PI{\V{S}}{\Set} \PIS{\V{S} \To \Desc} \Desc 
\end{array}
}
\vspace{0.2in}
\\
\stk{
\descop{\_\:}{} : \Desc \To \Set \To \Set \\
\begin{array}{@{}l@{\:=\:\:}ll}
\descop{\DId}{\V{X}}          &  \V{X}                                           \\
\descop{\DConst{\V{Z}}}{\V{X}}    &  \V{Z}                                           \\
\descop{\DProd{\V{D}}{\V{D'}}}{\V{X}} &  \TIMES{\descop{\V{D}}{\V{X}}}{\descop{\V{D}\V{'}}{\V{X}}}       \\
\descop{\DSigma{\V{S}}{\V{D}}}{\V{X}} &  \SIGMA{\V{s}}{\V{S}} \descop{\V{D}\: \V{s}}{\V{X}}          \\
\descop{\DPi{\V{S}}{\V{D}}}{\V{X}}    &  \PI{\V{s}}{\V{S}} \descop{\V{D}\: \V{s}}{\V{X}}            
\end{array}
}
\end{array}
}\]

@@

In the previous section, we have presented the $\Desc$ universe as the
grammar of functors in the category $\Set$. We have seen how to code
inductive types in this setting. To describe an inductive family
indexed by $\Bhab{\V{I}}{\Set}$, we use endofunctors on the category
$\Set^{\V{I}}$. We call these \emph{indexed functors}. \note{Predicate
  transformer anyone?} In this setting, $\V{I} \To \IDesc{\V{I}}$ is our
grammar for describing these functors. Hence, $\IDesc{}$ and its
interpretation have the following types:

\[\stk{
\data \IDesc{} (\Bhab{\V{I}}{\Set}) : \Set \where \\
\;\;\ldots \\
\\
\idescop{\_\:}{}{} : \PI{\V{I}}{\Set} \IDesc{\V{I}} \To (\V{I} \To \Set) \To \Set    \\
\ldots
}\]

Note that $I \To \IDesc{\V{I}}$ is isomorphic, up to interpretation, to an
endofunctor on $\Set^{\V{I}}$. Indeed, $\V{I} \To \IDesc{\V{I}}$ is interpreted as
$\V{I} \To (\V{I} \To \Set) \To \Set$, which is isomorphic to $(\V{I} \To \Set)
\To (\V{I} \To \Set)$. Fix-point of inductive families is defined over
these indexed functors, hence computing a fix-point of the entire
\emph{family} of functors:

\[\stkl{
\Rule{\stk{ 
          \Gamma \vdash \Bhab{\V{I}}{\Set} \\
          \Gamma \vdash \Bhab{\V{R}}{\V{I} \To \IDesc{\V{I}}} \\
          \Gamma \vdash \Bhab{\V{i}}{\V{I}}}}
     {\Gamma \vdash \Bhab{\IMu{\V{I}}{\V{R}}{\V{i}}}{\Set}} \qquad
\Rule{\stk{
          \Gamma \vdash \Bhab{\V{I}}{\Set} \\
          \Gamma \vdash \Bhab{\V{R}}{\V{I} \To \IDesc{\V{I}}} \\
          \Gamma \vdash \Bhab{\V{i}}{\V{I}} \\
          \Gamma \vdash \Bhab{\V{x}}{\idescop{\V{R}\: \V{i}}{\V{I}}{(\IMu{\V{I}}{\V{R}}{})}}}}
     {\Gamma \vdash \Bhab{\Con{\V{x}}}{\IMu{\V{I}}{\V{R}}{\V{i}}}}
}\]

@@

\[\stk{
\begin{array}{ll}
\stk{
\data \IDesc{} (\Bhab{\V{I}}{\Set}) : \Set \where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \DVar{}         & \V{I} \To \IDesc{\V{I}}                                   \\
    \DConst{}       & \Set \To \IDesc{\V{I}}                                \\
    \DProd{}{}      & \PI{\V{\V{D}}, \V{\V{D'}}}{\IDesc{\V{I}}} \IDesc{\V{I}}           \\
    \DSigma         & \PI{\V{S}}{\Set} \PIS{\V{S} \To \IDesc{\V{I}}} \IDesc{\V{I}}  \\
    \DPi            & \PI{\V{S}}{\Set} \PIS{\V{S} \To \IDesc{\V{I}}} \IDesc{\V{I}} 
\end{array}
}
\vspace{0.2in}
\\
\stk{
\idescop{\_\:}{}{} :_{\PI{\V{I}}{\Set}} \IDesc{\V{I}} \To (\V{\V{I}} \To \Set) \To \Set                  \\
\begin{array}{@{}l@{\:=\:\:}ll}
\idescop{\DVar{\V{i}}}{\V{I}}{\V{P}}      &  \V{P}~\V{i}                                                 \\
\idescop{\DConst{\V{X}}}{\V{I}}{\V{P}}    &  \V{X}                                                   \\
\idescop{\DProd{\V{D}}{\V{D'}}}{\V{I}}{\V{P}} &  \TIMES{\idescop{\V{D}}{\V{I}}{\V{P}}}{\idescop{\V{D'}}{\V{I}}{\V{P}}}       \\
\idescop{\DSigma{\V{S}}{\V{D}}}{\V{I}}{\V{P}} &  \SIGMA{\V{s}}{\V{S}} \idescop{\V{D}\: \V{s}}{\V{I}}{\V{P}}                    \\
\idescop{\DPi{\V{S}}{\V{D}}}{\V{I}}{\V{P}}    &  \PI{\V{s}}{\V{S}} \idescop{\V{D}\: \V{s}}{\V{I}}{\V{P}}            
\end{array}
}
\end{array}
}\]

@@

\[\stk{
 \TagIDesc{\V{I}}  \mapsto \TIMES{\ATagIDesc{\V{I}}}{\ITagIDesc{\V{I}}} \\
 \ATagIDesc{\V{I}} \mapsto \SIGMA{\V{E}}{\EnumU} (\PI{\V{i}}{\V{I}} \spi{\V{E}}{(\LAM{\_} \IDesc{\V{I}})}) \\
 \ITagIDesc{\V{I}} \mapsto \\
\qquad\SIGMA{\V{F}}{\V{I} \To \EnumU} (\PI{\V{i}}{\V{I}} \spi{(\V{F}\: \V{i})}{(\LAM{\_} \IDesc{\V{I}})}) 
}\]

@@

\[\stk{
\VecD : \Set \To \TagIDesc{\Nat} \\
\begin{array}{@{}ll}
\VecD\: \V{X} = \\ \quad\bigRedBracket{ 
            \begin{array}{l}
                \pair{[]}{\_}{} 
                \\\\
                \bigRedBracket{
                \begin{array}{l} 
                  \LAM{\V{n}} \caseB{\V{n}}{\NatZero \mapsto [\VecNil] \\
                                   (\NatSuc \V{n}) \mapsto [\VecCons{}{}]} \\
                  \LAM{\V{n}} \caseB{\V{n}}{\NatZero \mapsto [\DConst{\Unit}] \\
                                   (\NatSuc \V{n}) \mapsto [\DSigma{\V{X}}{\DVar{\V{n}}}]}
                \end{array}}
            \end{array}}
\end{array}
}\]

@@

\[\stk{
\IDescD : \PI{\V{I}}{\Set} \IDesc{\Unit} \\
\begin{array}{@{}ll}
\IDescD~\V{I} \mapsto \DSigma{}{} & (\EnumT \red{[} \DVar{},
                                          \DConst{},
                                          \DProd{}{},
                                          \DSigma{}{}, 
                                          \DPi{}{} \red{]}) \\
                              & \bigRedBracket{\begin{array}{l}
                                      \DConst{\V{I}}                  \\
                                      \DConst{\Set}               \\
                                      \DProd{\DVar{\Void}}{\DVar{\Void}}  \\
                                      \DSigma{\Set}{(\LAM{\V{S}} \DPi{\V{S}}{(\LAM{\_} \DVar{\Void})})} \\
                                      \DSigma{\Set}{(\LAM{\V{S}} \DPi{\V{S}}{(\LAM{\_} \DVar{\Void})})}
                                   \end{array}
                             }
\end{array}
}\]

@@

\[
\begin{array}{l@{\::\:\:}l}
\Econd{}{}{}     & \forall \Bhab{\V{ty}}{\Ty} . \Ebool \To \V{ty} \To \V{ty} \To \V{ty}  \\ 
\Eplus{}{}       & \Enat \To \Enat \To \Enat                           \\
\Ele{}{}         & \Enat \To \Enat \To \Ebool                          \\
\Eval{}          & \forall \Bhab{\V{ty}}{\Ty} . \Val{\V{ty}} \To \V{ty}
\end{array}
\]

@@

\[\stkc{
\stk{
\HExprD : \TagIDesc{\Ty} \\
\HExprD \mapsto ( \HExprAD , \HExprID ) \\
} \\
\\
\begin{array}{ll}
\stk{
\HExprAD : \ATagIDesc{\Ty} \\
\HExprAD \mapsto \bigRedBracket{
                 \begin{array}{l}
                   \EnumT{[\Eval{}, \Econd{}{}{}]} \red{,} \\
                   \LAM{\V{ty}}
                   \bigRedBracket{
                   \begin{array}{l}
                   \DConst{(\Val{\V{ty}})} \\
                   \DProd{\DVar{\Ebool}}{\DProd{\DVar{\V{ty}}}{\DVar{\V{ty}}}}
                   \end{array}
                   }
                 \end{array}
                 }
\\
\\
} &
\stk{
\HExprID : \ITagIDesc{\Ty} \\
\HExprID \mapsto \bigRedBracket{
                 \begin{array}{l}
                   \LAM{\V{t}y} \case{\V{t}y}{\Enat \To \EnumT{[\Eplus{\!}{\!}\!]} \\ \Ebool \To \EnumT{[\Ele{\!}{\!}\!]}} \red{,} \\
                   \LAM{\_} \DProd{\DVar{\Enat}}{\DVar{\Enat}} 
                   \end{array}
                   }
}
\end{array}
}\]

@@

\[
\evalH : \PI{\V{ty}}{\Ty} 
         \IMu{\Ty}{\toIDesc{\HExprD}}{\V{ty}} \To
         \Val{\V{ty}}
\]

@@

\[\stk{
\evalOne : \PI{\V{ty}}{\Ty} \idescop{\toIDesc{\HExprD}\: \V{ty}}{\Ty}{\V{ty}}
\To {\Val{\V{ty}}} \\
\begin{array}{@{}l@{\:=\:\:}l}
\evalOne\: \_\: \pair{\Eval{}}{\V{x}}{}                                         & \V{x} \\
\evalOne\: \_\: \pair{\Econd{}{}{}}{\pair{\BoolTrue}{\pair{\V{x}}{\_}{}}{}}{}   & \V{x} \\
\evalOne\: \_\: \pair{\Econd{}{}{}}{\pair{\BoolFalse}{\pair{\_}{\V{y}}{}}{}}{}  & \V{x} \\
\evalOne\: \Enat\: \pair{\Eplus{}{}}{\pair{\V{x}}{\V{y}}{}}{}                       & \F{plusHost}\: \V{x}\: \V{y} \\
\evalOne\: \Ebool\: \pair{\Ele{}{}}{\pair{\V{x}}{\V{y}}{}}{}                        & \F{leHost}\: \V{x}\: \V{y} 
\end{array} \\
\\
\evalH : \PI{\V{ty}}{\Ty} 
           \IMu{\Ty}{\toIDesc{\HExprD}}{\V{ty}} \To
           \Val{\V{ty}} \\
\evalH\: \V{ty}\: \V{term} = \F{cataI}\: \Ty\: 
                                 \toIDesc{\HExprD}\: 
                                 \Val{}\: 
                                 \evalOne\: 
                                 \V{ty}\: 
                                 \V{term}
}\]
\note{pwm: $\evalOne$ rather illustrates my point about the 2 types of
  IFunc floating around, $\idescop{\_}{}{}$ is defined using the 1st
  notion, and used it here as if it were the 2nd. pierre: in this
  case, it was a typo. But you're right, there is two notions and we
  probably want to clarify where these things live.}

@@

\[\stk{
\Var{}{} : \EnumU \To \Ty \To \Set \\
\Var{\V{dom}}{\_} = \EnumT{\V{dom}}
}\]

Whereas $\Val{}$ was mapping the type to the corresponding host type,
$\Var{}{}$ maps types to a finite set. The finite set -- the context
-- contains closed terms. A variable is therefore a $\Eval{}$ that
contains a pointer to a particular element of the finite set -- an
element of $\EnumT{\V{dom}}$. The extra argument to $\Var{}{}$ is the
domain of this context.

Consequently, replacing $\Val{\V{ty}}$ by
$(\SUM{\Val{\V{ty}}}{\Var{\V{dom}}{\V{ty}}})$ in Figure~\ref{fig:hexpr-full} turns
the language of closed term into a language of opened terms with
constants. For readability, we will abbreviate $\LAM{\V{ty}}
\SUM{\Val{\V{ty}}}{\Var{\V{dom}}{\V{ty}}}$ into $\SUM{\Val{}}{\Var{}{}}$ \note{Is
  it polite and comprehensible to ask that?} This defines a new
indexed description, called $\HExprVarD{dom}$.

@@

\[\stk{
\begin{array}{@{}ll}
\discharge : & \PITEL{\V{ty}}{\Ty}
               \PITEL{\V{dom}}{\EnumU} \\
             & \PI{\V{\gamma}}{\spi{\V{dom}}{(\LAM{\_} \IMu{\Ty}{\toIDesc{\HExprVarD{\V{dom}}}}{\V{ty}})}} \\
             & (\SUM{\Val{\V{ty}}}{\Var{\V{dom}}{\V{ty}}}) \To
               \IMu{\Ty}{\toIDesc{\HExprD}}{\V{ty}} 
\end{array} \\
\begin{array}{@{}l@{\:\mapsto\:\:}l}
\discharge\: \V{ty}\: \V{dom}\: \V{\gamma}\: (\SumLeft \V{x})  & \Con{\pair{\Eval{}}{\V{x}}{}} \\
\discharge\: \V{ty}\: \V{dom}\: \V{\gamma}\: (\SumRight \V{v}) &
\end{array}\\
\qquad\switch{\V{dom}}{(\LAM{\_}
\IMu{\Ty}{\toIDesc{\HExprVarD{\V{dom}}}}{\V{ty}})}{\V{\gamma}}{\V{v}}
}\]

@@

\newcommand{\substH}{\F{substExpr}}
\newcommand{\domNat}{dom_{\CN{nat}}}
\newcommand{\domBool}{dom_{\CN{bool}}}
\newcommand{\gammaNat}{\V{\ensuremath{\gamma_{\CN{nat}}}}}
\newcommand{\gammaBool}{\V{\ensuremath{\gamma_{\CN{bool}}}}}
\newcommand{\GammaSpi}[2]{\F{\ensuremath{\Gamma_{\CN{ty}}}}~#1~#2}

\[
\begin{array}{@{}ll}
\substH  : & \PITEL{\V{\dom}}{\EnumU} \\
           & \PITEL{\V{\gammaNat}}{\GammaSpi{\V{dom}}{\Enat}} 
             \PITEL{\V{\gammaBool}}{\GammaSpi{\V{dom}}{\Ebool}} \\
           & \begin{array}{@{}ll}
             \PI{\V{\sigma}}{& \PITEL{\V{dom}}{\EnumU} \\
                             & \PITEL{\V{\gammaNat}}
                                     {\GammaSpi{\V{dom}}{\Enat}} \\
                             & \PITEL{\V{\gammaBool}}
                                     {\GammaSpi{\V{dom}}{\Ebool}} \\
                             & \PI{\V{ty}}{\Ty} (\SUM{\Val{\V{ty}}}{\Var{\V{dom}}{\V{ty}}}) \To \\
                             & \IMu{\Ty}{(\toIDesc{\HExprD})}{\V{ty}}}
             \end{array}\\
          & \PI{\V{ty}}{\Ty}
            \IMu{\Ty}{\toIDesc{\HExprVarD{\V{dom}}}}{\V{ty}} \To \\
          & \IMu{\Ty}{\toIDesc{\HExprD}}{\V{ty}}
\end{array}
\]

@@

\[
\F{$\sigma$}\: \V{dom}\: \gammaNat\: \gammaBool\: \V{ty}\: \V{var} \mapsto
    \F{$\discharge$}\: \V{ty}\: \V{dom}\: \F{$\gamma$}_{\V{ty}}\: \V{var} 
\]
